\documentclass{article}

\usepackage{math}
\usepackage{algorithm, algpseudocode}

\title{Phase-field fracture models and MSPIN} % nb: placeholder
\author{Conor McCoid \and Blaise Bourdin}

\begin{document}

\maketitle

\begin{abstract}
---
\end{abstract}

\section{Introduction}
\label{sec: intro}

[What is brittle fracture]

[What is phase-field model]

[How has phase-field been computed in the past, with what limitations]

[Recent advances]

[Current proposition]

[Outline of paper]

\section{Background}
\label{sec: bg}

[Mathematical description of phase-field]

The phase-field energy may be expressed as
\begin{equation} \label{eq: energy} % nb: fix this
	E_\ell(\vec{u}, \vec{v}) = \int_{\Omega} \text{tensors} d\Omega + \int_{\partial \Omega} \left ( \frac{\omega(\vec{v})}{4 \ell} + \ell \abs{\vec{v}}^2 \right ) dS
	.
\end{equation}

[Irreversibility condition]

\subsection{Alternate Minimization}
\label{subsec: altmin}

[AltMin, using notation from mathematical description of phase-field]

AltMin may be viewed as a field-split multiplicative Schwarz method.
The two `fields' of damage and displacement are treated separately and sequentially.
This allows the use of some of the techniques reserved for Schwarz methods to be applied here.
In particular, in Section \ref{subsec: mspin} we will examine applying MSPIN to the problem.

\subsection{Active set methods}
\label{subsec: as}

\subsection{MSPIN}
\label{subsec: mspin}

Multiplicative Schwarz preconditioning inexact Newton (MSPIN) was first described in (nb: cite Cai and Keyes), with further explanation for field-split Schwarz methods in (nb: cite Liu and Keyes).
MSPIN has been previously applied to phase-field fracture models in (nb: cite 3K).
There, the authors used cubic backtracking to globalize the Newton direction, see Section \ref{subsec: cb} below.
Note that their model differs slightly from the one presented here.

We present MSPIN as applied to AltMin.
[fill in once notation is specified from prev. section]

\subsection{Cubic Backtracking}
\label{subsec: cb}

Many applications of Newton's method require globalization techniques to achieve convergence when starting away from the minimum.
When the Newton direction overshoots the minimum, these globalization techniques should search for a minimum between the starting point and the destination, usually through a line search.

Cubic backtracking (nb: cite) performs this line search by repeatedly interpolating cubic polynomials and finding their minima.
Note that a cubic polynomial interpolant requires four pieces of information, but Newton's method only provides three:
the value of the function at the starting point, $f(\vec{x})$;
the value of the function in the Newton direction, $f(\vec{x} + \vec{p})$, and;
the derivative of the function in the Newton direction, $\vec{p} \cdot \nabla f(\vec{x})$.

To get a fourth piece of information, first form a quadratic interpolant with this information.
Find the minimum of this polynomial and test the value of the function there, thus getting enough information for a cubic polynomial.
If the function is smaller at this point, then this ends the line search.
Otherwise, find the minimum of the cubic interpolant and repeat.
This algorithm is summarized in Algorithm \ref{alg: cb}.
\begin{algorithm} \label{alg: cb}
	\caption{Cubic backtracking (nb: cite)}
	\begin{algorithmic}[1]
	\State inputs
	\end{algorithmic}
\end{algorithm}

\section{Interpolant minimization}

When using MSPIN, one has two potential directions for minimization: the AltMin direction $\vec{q}$ and the Newton direction $\vec{p}$.
Let's try using both while applying similar ideas to cubic backtracking.
Rather than find the minimum of a single variable cubic polynomial, find the minimum of a multivariate quadratic polynomial interpolant.
Here we present three options for interpolants to minimize.

\subsection{Parallelogram}
\label{subsec: para}

We begin by considering a 2D interpolant using the AltMin and Newton directions, $\vec{q}$ and $\vec{p}$.
For this interpolant, we need six pieces of information.
Five are provided by MSPIN:
the value of the function at the starting point, $f(\vec{x})$;
the value of the function in the AltMin direction, $f(\vec{x} + \vec{q})$;
the value of the function in the Newton direction, $f(\vec{x} + \vec{p})$;
the derivative of the function in the AltMin direction, $\vec{q} \cdot \nabla f(\vec{x})$, and;
the derivative of the function in the Newton direciton, $\vec{p} \cdot \nabla f(\vec{x})$.

One choice for a sixth piece of information is the value of the function at a point symmetric to the starting point.
Suppose that the AltMin and Newton directions form two sides of a parallelogram.
Then the other corners of the parallelogram are the starting point and $\vec{x} + \vec{p} + \vec{q}$.
It is at this new point that we evaluate the function and find our sixth piece of information.

We seek a point $\vec{x} + \alpha \vec{q} + \beta \vec{p}$ that minimizes the objective function.
The interpolant appears as $P(\alpha,\beta) = a \alpha^2 + b \alpha \beta + c \beta^2 + d \alpha + e \beta + f$.
It is straightforward to show how our six pieces of information relate to our six coefficients:
\begin{align*}
	d = \vec{q} \cdot \nabla f(\vec{x}), \qquad
	e = \vec{p} \cdot \nabla f(\vec{x}), \qquad
	f = f(\vec{x}), \\
	a = f(\vec{x} + \vec{q}) - d - f, \qquad
	c = f(\vec{x} + \vec{p}) - e - f, \\
	b= f(\vec{x} + \vec{q} + \vec{p}) + f - f(\vec{x} + \vec{q}) - f(\vec{x} + \vec{p}).
\end{align*}
The minimum of this interpolant occurs at
\begin{equation*}
	\alpha = \frac{be - 2cd}{4ac-b^2}, \quad \beta = \frac{-db - 2ae}{4ac-b^2}.
\end{equation*}

It's possible for the minimum of the polynomial to occur far outside of the parallelogram.
This may be acceptable, but could also destabilize convergence.
To stay within the parallelogram, we can also check the minima on the boundary:
\begin{align*}
	(0, -e/2c), && (-d/2a, 0), && (1, -(e+b)/2c), && (-(d+b)/2c, 1).
\end{align*}
In practice, it is advisable to have $\beta$, the amount to step in the Newton direction, be between -1 and 1,
while $\alpha$, the amount to step in the AltMin direction, can be any positive value.
We will see examples of this behaviour in the next section.

% nb: what about the smallest value within a tolerated distance of the starting point?

Before accepting the critical point of the interpolant as the next step in the algorithm, one should perform the second derivative test to confirm that this point is in fact a minimum.
The determinant of the Jacobian matrix is $D = 4ac - b^2$, used in the denominator of both $\alpha$ and $\beta$.
If $D>0$ and $a>0$, then the critical point is indeed a minimum.
Otherwise, this point may be a maximum or a saddle point.
This becomes more likely when the calculation of the gradient of the function is subject to numerical error, especially when the magnitude of the gradient is small in the AltMin and Newton directions.

Should the calculated critical point be deemed unacceptable, either by being too far from the starting point or failing the second derivative test, then one should change strategies.
One can choose one of the other interpolants suggested here, or select one of the three corners of the parallelogram with the lowest energy as the next step.

\subsection{Triangle}
\label{subsec: tri}

If one wants to form interpolants using only evaluations of the objective function and not its gradient, then one can form a P2 finite element by evaluating the objective function at six points.
Four of these points are the corners of the parallelogram described previously.
The remaining two are $\vec{x} + 2 \vec{p}$ and $\vec{x} + 2 \vec{q}$.
The coefficients of the polynomial $P(\alpha, \beta)$ are then:
\begin{align*}
	f = f(\vec{x}), \qquad
	b = f(\vec{x} + \vec{q} + \vec{p}) + f - f(\vec{x} + \vec{q}) - f(\vec{x} + \vec{p}), \\
	a = \frac{1}{2} \left ( f(\vec{x} + 2 \vec{q}) + f \right ) - f(\vec{x} + \vec{q}), \qquad
	d = f(\vec{x} + \vec{q}) - f - a, \\
	c = \frac{1}{2} \left ( f(\vec{x} + 2 \vec{p}) + f \right ) - f(\vec{x} + \vec{p}), \qquad
	e = f(\vec{x} + \vec{p}) - f - c.
\end{align*}
Like the parallelogram, the minimum of this interpolant occurs at
\begin{equation*}
	\alpha = \frac{be - 2cd}{4ac-b^2}, \quad \beta = \frac{-db - 2ae}{4ac-b^2}.
\end{equation*}

If one started with the parallelogram interpolant but were unsatisfied by its critical point, then the triangle interpolant requires only two more evaluations of the objective function.
If the new interpolant still fails the second derivative test, then the next step in the algorithm can be chosen as that which has the lowest value of the objective function from amongst the five values already found.

\subsection{Tetrahedron}
\label{subsec: tet}

Lastly, let us consider forming a 3D interpolant by including the intermediate step in AltMin.
That is, let $\vec{p}$ and $\vec{q}$ remain the same, and add in the direction $\vec{r}$, which is the restriction of $\vec{q}$ to the damage degrees of freedom.

To form the P2 finite element for this 3D interpolant, we need 10 evaluations of the objective function.
This will be the six already used for the triangle interpolant, as well as $f(\vec{x} + \vec{r})$, $f(\vec{x} + \vec{r} + \vec{q})$, $f(\vec{x} + \vec{r} + \vec{p})$, and $f(\vec{x} + 2 \vec{r})$.
The interpolant now appears as
\begin{equation*}
	Q(\alpha, \beta, \gamma) = a \alpha^2 + b \beta^2 + c \gamma^2 + d \alpha \beta + e \beta \gamma + f \alpha \gamma + g \alpha + h \beta + i \gamma + j.
\end{equation*}
These coefficients are equal to
\begin{align*}
	j = f(\vec{x}), \qquad
	d = f(\vec{x} + \vec{p} + \vec{q}) - f(\vec{x} + \vec{q}) - f(\vec{x} + \vec{p}) + j, \\
	e = f(\vec{x} + \vec{r} + \vec{p}) - f(\vec{x} + \vec{p}) - f(\vec{x} + \vec{r}) + j, \\
	f = f(\vec{x} + \vec{q} + \vec{r}) - f(\vec{x} + \vec{r}) - f(\vec{x} + \vec{q}) + j,\\
	a = \frac{1}{2} \left ( f(\vec{x} + 2 \vec{q}) - 2 f(\vec{x} + \vec{q}) + j \right ), \qquad
	g = f(\vec{x} + \vec{q}) - j - a, \\
	b = \frac{1}{2} \left ( f(\vec{x} + 2 \vec{p}) - 2 f(\vec{x} + \vec{p}) + j \right ), \qquad
	h = f(\vec{x} + \vec{p}) - j - b, \\
	c = \frac{1}{2} \left ( f(\vec{x} + 2 \vec{r}) - 2 f(\vec{x} + \vec{r}) + j \right ), \qquad
	i = f(\vec{x} + \vec{r}) - j - c.
\end{align*}
The minimum occurs at the solution to the following system:
\begin{equation*}
	\begin{bmatrix} 2a & d & f \\ d & 2b & e \\ f & e & 2c \end{bmatrix} \begin{bmatrix} \alpha \\ \beta \\ \gamma \end{bmatrix} = - \begin{bmatrix} g \\ h \\ i \end{bmatrix},
\end{equation*}
but only if the matrix of this system is positive definite.
Otherwise, this critical point is a saddle point or a maximum.

\section{Numerical comparisons}
\label{sec: num}

Numerical experiments conducted for this article are run using FEniCSx for its finite element library, petsc4py for solving linear systems, and mpi4py for parallelization.
The Docker image used for these experiments can be found at (nb: DockerHub link) (nb: cite Docker and Apptainer).
Code for these experiments can be found at (nb: GitHub link).

\subsection{Surfing boundary conditions}
\label{subsec: surf}

For this example, we begin with an initial crack of length (nb: length) in an otherwise rectangular domain with an unstructured grid of (nb: number) points.
The boundary conditions set the damage at maximum on the initial crack and zero on the boundary.
For displacement, the boundary conditions are of Neumann type, setting
\begin{align*}
	\frac{\partial u}{\partial x} = \frac{K}{2 \mu} \sqrt{\frac{r(t)}{2 \pi}} \left ( k - \cos \theta(t) \right ) \cos \frac{\theta(t)}{2},
	\quad
	\frac{\partial u}{\partial y} = \frac{K}{2 \mu} \sqrt{\frac{r(t)}{2 \pi}} \left ( k - \cos \theta(t) \right ) \sin \frac{\theta(t)}{2},
	\\
	r(t) = \sqrt{(x-t)^2 + y^2}, \quad \theta(t) = \arctan \frac{y}{x-t},
\end{align*}
where $t$ defines the force exerted on the system, (nb: and define the remaining parameters).

[figure of domain, crack progression]

[compare energy]

[compare outer iterations]

[discussion on number of inner iterations]

[compare flop counts]

[compare wall clock time]

\subsection{CTFM (nb: expand acronym)}
\label{subsec: ctfm}

\end{document}